---
title: "Plotting spatial residuals"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plotting spatial residuals}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

Here we will use the vista package to display spatial residuals from spatiotemporal models

```{r set-knitr-options, cache=FALSE, echo=FALSE, warning=FALSE, message=FALSE}
library("knitr")
opts_chunk$set(message = FALSE, fig.width = 5.5)
```

```{r}
library(vista)
library(mgcv)
data(pcod)
```

### GAM with spatial smooths by year

We'll use a GAM for demonstration purposes, but any other package `randomForest`, `sdmTMB`, etc. could be used instead.  

```{r}
m <- gam(density ~ 0 + as.factor(year) + s(X,Y,by=as.factor(year)), 
         data=pcod, family=tw())
pcod$pred = predict(m)
pcod$resid = residuals(m)
```

### Predictions in space

The `pred_space` function plots predictions spatially. By default, removes temporal means so the predictions in each time slice have mean 0. By default, the predictions are faceted by time,  
```{r}
pred_space(df = pcod, time = "year")
```

But we can also turn off faceting using `by_time`  
```{r}
pred_space(df = pcod, time = "year", by_time = FALSE)
```

### Residuals  in space

Similarly, the `resid_space` function is designed to plot spatial residuals (again these have mean 0). And faceting can be turned on / off, but by default `by_time = TRUE`. Asterisks are used to indicate outliers, or values of residuals that are more than 3 standard deviations in magnitude.

```{r}
resid_space(df = pcod, time = "year")
```

## QQ plot spatially

The `qq_space` function is also useful in displaying the quantile residuals (via a call to `stats::qqnorm`)  

```{r}
qq_space(pcod, time = "year")
```

