---
title: "Plotting temporal residuals"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plotting temporal residuals}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

Here we will use the vista package to plot temporal residuals from spatiotemporal models

```{r set-knitr-options, cache=FALSE, echo=FALSE, warning=FALSE, message=FALSE}
library("knitr")
opts_chunk$set(message = FALSE, fig.width = 5.5)
```

```{r}
library(vista)
library(mgcv)
data(pcod)
```

### GAM with spatial smooths by year

We'll use a GAM for demonstration purposes, but any other package `randomForest`, `sdmTMB`, etc. could be used instead.  

```{r}
m <- gam(density ~ 0 + as.factor(year) + s(X,Y,by=as.factor(year)), 
         data=pcod, family=tw())
pcod$pred = predict(m)
pcod$resid = residuals(m)
```

### Predictions through time

The `pred_time` function displays the range of predictions across time slices. 
```{r}
pred_time(df = pcod, time = "year")
```

### Residuals through time

Similarly, the `resid_time` function is designed to plot residuals across time slices  

```{r}
resid_time(df = pcod, time = "year")
```

### Homogeneity of residuals through time

In some cases, the variability of a spatial process may be non-stationary and increasing or decreasing over time. We can visualize this using the 
`sd_resid_time` function, where ideally there is no trend or outlier.  

```{r}
sd_resid_time(df = pcod, time = "year")
```

## QQ plot temporally

The `qq` function is also useful in displaying the quantile residuals (via a call to `stats::qqnorm`). By default this will be faceted by time, but faceting can be turned off with `by_time = FALSE`. 

```{r}
qq(pcod, time = "year")
```

## Scale - location plots

The `scale_loc` function displays scale - location plots, which may be useful at diagnosing trends in residuals vs predicted values,  

```{r}
scale_loc(df = pcod, time="year")
```


## Moran's I statistics

We can calculate lots of variants of Moran's I. The `moran_ts` plot is useful for displaying time series of the Moran-I statistic, calculated separately for each time slice. This can be done on predicted values, 

```{r}
moran_ts(df = pcod, time = "year", response = "pred")
```  

But if a goal is evaluating whether a particular model removes spatial autocorrelation, it may be more useful to calculate this on the residuals,  

```{r}
moran_ts(df = pcod, time = "year", response = "resid")
```  


